Preprocessing and exploration of the narwhal data for the analysis with smoothSDE

### 2024-11-08 R/preprocessing/narwhals/data_preprocessing.Rmd
reorganize folders

@@

plots_folder <-"plots"

@@

```{r read raw data,cache=TRUE}
# Set the path to the directory containing the data
par_dir=dirname(dirname(getwd())) #parent directory 
narwhal_data_path <- file.path(par_dir,"Data", "Narwhals")

#get narwhal data
data_file="db_narwhal_2017_2018_with_gps.txt"
rawData <- read.table(file.path(narwhal_data_path,data_file),
                      header = TRUE, sep = "\t",dec = ".")

@@

```{=tex}
Two controlled exposure experiments were performed in 2017 and 2018.
The main differences are that only small guns were used in 2017 while only big guns were used in 2018, and the tagged whales used a much larger part of Scoresby Sound in 2018.

First preprocessing steps include : 
\begin{itemize}
\item keeping only 2018 data - adding a time since tagging column 
\item converting distance to ship and distance to shore to km - adding an ExpShip column defined as the inverse of the \texttt{Dist\_Ship} column when the narwhals are in line of sight from the boat.
\item Renaming the "Ind" column to "ID" and converting it to factors.
\item Converting column "Type" in factor with levels trial and intertrial 
\item Converting column "Posixct" to POSIXct format for date and time
\end{itemize}

The first measurement time is set to $1$ second and then measures are interpolated every second (see Jorgensen et al.).
Thus, the time $T_k$ of the $k$-th measurement is $$T_{k}=\frac{k}{3600}\mbox{ h }$$ As proposed in Jorgensen et al., we also add a column with a covariate $E$ defined as $$E(t)=\frac{1}{\mbox{distance to ship}(t)}$$ where distance to ship has been converted in km.

\texttt{ExpShip} is a proxy for sound exposure level (SEL).
```

@@

# Data exploration after first preprocessing

@@

#plot with all the narwhals rescaled so that the experiments start at the same time 
trials1=ggplot(data[!is.na(data$Type), ], aes(time/24,ID,fill=Type)) + 
  xlab("Number of days after tagging")+ylab("Exposure")+
  geom_tile(lwd = 1,linetype = 1)+ 
  scale_fill_manual(values=c("blue","red"),na.value="white",
                    name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

#plot with all the narwhals at the real experiments time
trials2=ggplot(data[!is.na(data$Type), ], aes(Posixct,ID,fill=Type)) + 
  xlab("Day")+ylab("Exposure")+
  geom_tile(lwd = 1.5,linetype = 1)+ 
   scale_fill_manual(values=c("blue","red"), na.value="white",
                     name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

ggsave(path=plots_folder,filename="trials_rescaled.png",plot=trials1,width=10,height=5)
ggsave(path=plots_folder,filename="trials.png",plot=trials2,width=10,height=5)
```

## Ship exposure through time and depending on trial type

```{r define ExpShipI and ExpShipT}

It appears that the exposure is not so smooth with respect to time, meaning that a narwhal can be highly exposed at some time and be completely out of sight from the airguns a few seconds later.
This phenomenon appears when the values in DistShip suddenly change to NA.

Moreover, exposure times are relatively short compare to the overall duration of the measurement period, so that most of the time, ExpShip is zero but short peaks appear when trials or intertrials are happening.
We would need to zoom on these peaks to assess their smoothness.

@@

#plot exposure durinh intertrial
pExpInt=ggplot(data,aes(time,ExpShipI,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

@@

```{=tex}
\textbf{\texttt{Lat} and \texttt{Long} columns have been interpolated every second, and distance to ship has been computed from these interpolations.
The true measured distance to ship are in the rows for which the column \texttt{pos\_org}  has value 1.}
```

@@

```{=tex}
\begin{itemize}
\item  Drop missing latitude and longitude
\item Keep GPS positions indicated by value 1 in \texttt{pos\_org} column
\item  Project to UTM zone 26. The narwhal positions are in latitude and longitude.To analyze the track and visualize it better we project it on a plane in UTM coordinastes.
The positions of the narwhals are in the UTM zone 26 north (in Greenland).
\item Add a column "speed" with empirical horizontal speed obtained from the GPS positions (not interpolated)
\end{itemize}

@@

#remove rows with unknown positions
dataprep=data[!(is.na(data$Lat)) & !(is.na(data$Long)),]

@@

#compute empirical horizontal speed
n=length(dataprep$time)
dataprep$horizontal_speed=rep(NA,n)

@@

## measurement periods for each narwhal
```{r measurement periods}
#print measurement periods for each narwhal

@@

```{=tex}
The mean speed is about 4.5 km/h.
Usually, computing the empirical horizontal speed between two positions amounts to consider that the narwhal moved in a straight line between
the two points. But this is obviously not true, especially for long time steps, because the narwhal make U-turns for instance, and more importantly, it dives to find food. Therefore the empirical horizontal speed always underestimates the true horizontal speed.
Anormal values of the speed may indicate outliers to remove from the data.  Here we clearly see some outliers in the histogram
```

@@

@@

# Time steps distribution

@@

# Time steps distribution during exposure vs no exposure

# Plot all GPS tracks (without interpolation) on a map

@@

```{r save points as geojson in data folder}
points_sf <- st_as_sf(dataprep[,c("x","y")], coords = c("x","y"))
st_crs(points_sf)="+init=EPSG:32626 +units=km"
st_write(points_sf,file.path(narwhal_data_path,"narwhals_positions.shp"),append=FALSE)
```

# Coastline and land data

@@

```{r extract coastline and land}

# Set the path to the directory containing coastline data
par_dir=dirname(dirname(getwd())) #parent directory 
greenland_data_path <- file.path(par_dir,"Data", "Greenland")

#get land polygons
land_utm<-st_read(file.path(greenland_data_path,"land_utm.shp"))

@@

## Plot coastline and land

```{r plot coastline and land}
plot_land=ggplot()+geom_sf(data=land_utm$geometry,fill="grey",col="brown")+
  coord_sf(datum=st_crs(32626))
```

@@

## Compute nearest point on land and coastline for each GPS position

@@

nearest_points=function(positions,coastline) {
  #positions: dataframe of positions in UTM coordinates in metres
  #coastline : geometry of the coastline projected in the same UTM zone as positions in metres

@@

    # Iterate over each polygon in coastline
    for (j in 1:length(coastline$geometry)) {

      # Get the current polygon
      polygon <- coastline$geometry[[j]]

@@

#nearest points in coastline and land

@@

res_land=nearest_points(positions=narwhal_positions,coastline=land_utm)
})

@@

## Compare computed distances to shore

@@

For the rest of the analysis, we'll keep the bounded domain defined by the polygons in \texttt{modified\_land\_utm} sf object as a reference.
Doing that involves having some points on land, that is points with Distance to shore equal to 0.
This should be essentially due to the error in the land polygons than to the error in GPS measurements.
```

@@

For coastline, distance is 0 only if the position of the narwhal is exactly on the coastline or on a small island. A proper "land domain" is not defined in this case, but only the boundary of the domain (the coastline).

@@

cat("Percentage of points on land with distance from land : \n",
    length(which(res_land$min_distances==0))/length(dataprep$time)*100,"\n")

@@

```{r check original points on land}
check_plot1=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=dataprep,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

ggplotly(check_plot1)
```

```{r inspect points on land}
dataprep[dataprep$DistanceShore==0,c("Depth","time")]
```

```{r check new points on land}
df=dataprep
df$DistanceShore=res_land$min_distances

check_plot2=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

ggplotly(check_plot2)
```

```{r percentage points on land per ID}

@@

@@

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<1 & res_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")

@@

This seems to be pretty uniformly distributed. It is difficult to assess at which distance the narwhals stay from 
the shore in reality.

@@

#keep points calculated from land polygons
nearest_points=res_land$nearest_points
min_distances=res_land$min_distances

@@

```{r}
#add ExpShore covariate based on 1/distance in km
dataprep$ExpShore <- ExposureFunction(X = dataprep$DistanceShore)

@@

```{=tex}
It turns out that the effect was due to points on land. 
We can now observe a slight decrease in the angle of deviation from the coastline, which may indicate that narwhals tend to move parallel to the coast when close enough.
However, no effects of Distance to shore is observed on the empirical deviation between consecutive velocities.
This may be due to the averaging that is made to get those plots. The narwhal only turns for a few positions and then keep its heading paralell to the shore.
```

@@

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=sub_data,aes(px,py),shape=5,size=0.1,col="red")+
  geom_point(data=sub_data,aes(x,y,col=time),size=0.1)+
    geom_path(data=sub_data,mapping=aes(x,y),linewidth=0.1)+
    geom_point(data = sub_data %>% filter(!duplicated(ID)),
             aes(x,y,col=ID), shape = 3, size = 4,col = "red")+
    scale_color_viridis_c()

@@

```{r display Dship_histo, fig.cap="Histograms of distance to ship"}
Dship_histo
```
# Further details

## Plot ship exposure without interpolated points

```{r plot real exposure}
pExpShip=ggplot(dataprep,aes(time,ExpShip,col=ID))+geom_line(color="black")+
  facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to ship")

```{r}
df=dataprep
df[df$DistanceShore<D_low,"ExpShore"]=1/D_low
df[df$DistanceShore>D_up,"ExpShore"]=0

pExpShore=ggplot(df,aes(time,ExpShore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to shore")

ggsave(path=plots_folder,"realExpShore_through_time.png", plot=pExpShore)
```

```{r}
pExpShore
```

@@

dataBE1=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE1 =rbind(dataBE1,subset(narwhalData, !(time > maxtime) & (time>tagtime)))
}

dataBE2=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE2 =rbind(dataBE2,subset(narwhalData, !(time > maxtime) & (time>tagtime/2)))
}

dataAE=data.frame()
for (id in unique(dataprep$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataAE =rbind(dataAE,subset(narwhalData, (time > maxtime) & (time>tagtime)))
}

@@

```{r save preprocessed data}

write.csv(dataprep,file.path(narwhal_data_path,"allData.csv"))
write.csv(dataBE1,file.path(narwhal_data_path,"DataBE1.csv"))
write.csv(dataBE2,file.path(narwhal_data_path,"DataBE2.csv"))
write.csv(dataAE,file.path(narwhal_data_path,"DataAE.csv"))
```

>>>>

#to get root of git repo
library(here)

@@

plots_folder <-"exploratory_plots"

@@

```{r read raw data,cache=TRUE}
# Set the path to the directory containing the raw data for narwhals
raw_narwhal_data_path <- here(file.path("Data","raw_data","narwhals"))

#get narwhal data
data_file="db_narwhal_2017_2018_with_gps.txt"
rawData <- read.table(file.path(raw_narwhal_data_path,data_file),
                      header = TRUE, sep = "\t",dec = ".")

@@

```{=tex}
Two controlled exposure experiments were performed in 2017 and 2018.
The main differences are that only small guns were used in 2017 while only big guns were used in 2018, and the tagged whales used a much larger part of Scoresby Sound in 2018. More interesting results were found in 2018 in the paper of Mads-Peter, Adeline and Susanne. We primarily focus on this data rather than the 2017 data.

First preprocessing steps include : 
\begin{itemize}
\item keeping only 2018 data and adding a time since tagging column 
\item converting distance to ship and distance to shore to km and adding an ExpShip column defined as the inverse of the \texttt{Dist\_Ship} column when the narwhals are in line of sight from the boat.\texttt{ExpShip} is a proxy for sound exposure level (SEL).
\item Renaming the "Ind" column to "ID" for clarity and converting it to factors.
\item Converting column "Type" in factor with levels trial and intertrial, and adding two columns "ExpShipT" and "ExpShipI" for exposure to shio during trial and intertrial separately
\item Converting column "Posixct" to POSIXct format for date and time
\end{itemize}

The first measurement time is set to $1$ second and then measures are interpolated every second (see Jorgensen et al.).
Thus, the time $T_k$ of the $k$-th measurement is $$T_{k}=\frac{k}{3600}\mbox{ h }$.

@@

# First data exploration

@@

#plot with all the narwhals rescaled so that the experiments start at the same time 
trials_rescaled=ggplot(data[!is.na(data$Type), ], aes(time/24,ID,fill=Type)) + 
  xlab("Number of days after tagging")+ylab("Exposure")+
  geom_tile(lwd = 1,linetype = 1)+ 
  scale_fill_manual(values=c("blue","red"),na.value="white",
                    name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

#plot with all the narwhals at the real experiments time
trials=ggplot(data[!is.na(data$Type), ], aes(Posixct,ID,fill=Type)) + 
  xlab("Day")+ylab("Exposure")+
  geom_tile(lwd = 1.5,linetype = 1)+ 
   scale_fill_manual(values=c("blue","red"), na.value="white",
                     name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

ggsave(path=plots_folder,filename="trials_rescaled.png",plot=trials_rescaled,width=10,height=5)
ggsave(path=plots_folder,filename="trials.png",plot=trials,width=10,height=5)
```

The exposure is not smooth with respect to time, because a narwhal can be highly exposed at some time and be completely out of sight from the airguns a few seconds later, and also because the airguns shooting stops abruptly.
This phenomenon appears when the values in DistShip suddenly change to NA.

Moreover, exposure times are relatively short compare to the overall duration of the measurement period, so that most of the time, ExpShip is zero but short peaks appear when trials or intertrials are happening.
We can zoom on these peaks for a better visualization.

@@

#plot exposure during intertrial
pExpInt=ggplot(data,aes(time,ExpShipI,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

@@

@@

```{=tex}
\textbf{\texttt{Lat} and \texttt{Long} columns have been interpolated every second, and distance to ship has been computed from these interpolations.The true measured distance to ship are in the rows for which the column \texttt{pos\_org}  has value 1.}
We will drop the interpolated positions for the analysis.
```

```{=tex}
The next step consists in :
\begin{itemize}
\item  Dropping missing latitude and longitude.
\item Keeping only GPS positions indicated by value 1 in \texttt{pos\_org} column.
\item  The narwhal positions are in latitude and longitude. To analyze the track and visualize it better we project it on a plane in UTM coordinastes. Hence, we project to UTM zone north 26 (https://epsg.io/32626).
\item We add a column "speed" with empirical horizontal speed obtained from the GPS positions (not interpolated)
\end{itemize}

@@

#remove rows with missing positions
dataprep=data[!(is.na(data$Lat)) & !(is.na(data$Long)),]

@@

#compute empirical horizontal speed
dataprep$horizontal_speed=rep(NA,nrows)

@@

```{=tex}
We plot exposure from interpolated positions and highlight points that are directly observed exposures, that is exposure levels computed from the actual GPS position of the narhwal compared to the GPS position of the ship.

## Measurement periods for each narwhal

```{r measurement periods}
#print measurement periods for each narwhal

@@

```{=tex}
The mean speed is about 4.5 km/h.
Usually, computing the empirical horizontal speed between two positions amounts to consider that the narwhal moved in a straight line between
the two points. But this is obviously not true, especially for long time steps, because the narwhal make U-turns for instance. Therefore the empirical horizontal speed always underestimates the true horizontal speed.
Anormal values of the speed may indicate outliers to remove from the data.  While there is no reason to drop low speed values, since the narwhals can well rest for some time, high speed values that are clearly outliers and could not have been reached by a narwhal might be removed from the data.
Here we clearly see some outliers in the histogram.
```

@@

```{=tex}
Anormal values of the speed may indicate outliers to remove from the data.  Here we clearly see some outliers in the histogram.
We thus decide to filter by setting a threshold of $20$ km/h for horizontal speed.
```

@@

```{r histogram filtrered horizontal speed,warnings=FALSE}
speed_histo_filtered=ggplot(dataprep, aes(x =horizontal_speed,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")

speed_histo_filtered
ggsave(path=plots_folder,"speed_histo_filtered.png", plot = speed_histo_filtered)
```
## Time steps distribution

```{=tex}
Now that we removed the interpolated positions, we can check the time step distribution to get the frequence of acquisition of the narwhals Fastloc GPS device.
```

@@

## Time steps distribution during exposure vs no exposure

## Plot all GPS tracks (without interpolation) on a map

@@

```{r save points as shapefile in data folder}
# Set the path to the directory containing the preprocessed data
preprocessed_narwhal_data_path <- here(file.path("Data","preprocessed_data","narwhals"))

points_sf <- st_as_sf(dataprep[,c("x","y")], coords = c("x","y"))
st_crs(points_sf)="+init=EPSG:32626 +units=km"
st_write(points_sf,file.path(preprocessed_narwhal_data_path,"narwhals_positions.shp"),append=FALSE)
```

# Land polygons data for the domain boundary

@@

We read the preprocessed land polygons extracted from OpenStreetMap database.

```{r extract land polygons}

# Set the path to the directory containing coastline data
preprocessed_greenland_data_path <- here(file.path("Data","preprocessed_data", "greenland"))

#get land polygons
scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"scoresby_sound_utm.shp"))

@@

## Plot scoresby sound land

```{r plot coastline and land}
plot_land=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey",col="brown")+
  coord_sf(datum=st_crs(32626))
```

@@

## Compute nearest point on land for each GPS position

@@

nearest_points=function(positions,border) {
  #positions: dataframe of positions in UTM coordinates in metres
  #border : geometry of the coastline projected in the same UTM zone as positions in metres

@@

    # Iterate over each polygon in border
    for (j in 1:length(border$geometry)) {

      # Get the current polygon
      polygon <- border$geometry[[j]]

@@

#nearest points in border and land

@@

res_land=nearest_points(positions=narwhal_positions,border=scoresby_sound_utm)
})

@@

## Compare distances to shore from original data and computed distances to shore

@@

@@

@@

cat("Percentage of points on land with distance from osm land polygons : \n",
    length(which(res_land$min_distances==0))/length(dataprep$time)*100,"\n")

@@

```{r check original points on land}
plot_original_inland_pts=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=dataprep,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

ggplotly(plot_original_inland_pts)
```

```{r inspect depth of points on land}
dataprep[dataprep$DistanceShore==0,c("Depth","time")]
```

```{r check osm points on land}
df=dataprep
df$DistanceShore=res_land$min_distances

plot_osm_inland_pts=ggplot()+geom_sf(data=scoresby_sound_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

ggplotly(plot_osm_inland_pts)
```

```{r percentage points on land per ID with osm map}

@@

## Use land polygons updated with QGIS

```{=tex}
We need to know the nearest point on land to compute how the narwhal moves with respect to the land. For this, we need use the land polygons. In the original data, we have values for the Distance to shore, but we don't know exactly how it was computed. Approximately $2 \%$ of the narwhals positions are inland. This may be due to GPS measurement error, or errors in distance to shore computations. We notice that some points that have non zero distance to shore in the original data appear inland on the osm map. This suggests the osm map is not accurate. Indeed, as we have just seen, if we compute the distance to shore based on the osm polygon data, $9\%$ of the narwhal positions are inland. Thus, we decided to manually inspect the polygon data from OpenStreetMap with qgis software. We overlayed Sentinel-2 satellite images with the osm polygons data and the narwhals positions, and modified the shoreline to better match the original distance to shore and the satellite image. The resulting shapefile is updated_scoresby_sound_utm.shp.
```

```{r extract updated land polygons}

#get updated land polygons
updated_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"updated_scoresby_sound_utm.shp"))

updated_scoresby_sound_utm <- st_transform(updated_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")

```{r compute updated distance to shore}
res_updated_land=nearest_points(positions=narwhal_positions,border=updated_scoresby_sound_utm)

cat("Percentage of points on land with distance from updated osm land polygons : \n",
    length(which(res_updated_land$min_distances==0))/length(dataprep$time)*100,"\n")

```{r check updated osm points on land}
df=dataprep
df$DistanceShore=res_updated_land$min_distances

plot_updated_osm_inland_pts=ggplot()+geom_sf(data=updated_scoresby_sound_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

ggplotly(plot_updated_osm_inland_pts)
```

@@

df=data.frame(DistanceShore=res_updated_land$min_distances[res_updated_land$min_distances<1 & res_updated_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")

@@

For the rest of the analysis, we'll keep the bounded domain defined by the updated polygons in "updated_scoresby_sound_utm" sf object as a reference.
Doing that involves having some points on land, that is points with Distance to shore equal to 0.
```

@@

#keep points calculated from land polygons
nearest_points=res_updated_land$nearest_points
min_distances=res_updated_land$min_distances

@@

```{r}
#add Ishore covariate based on 1/distance in km
dataprep$Ishore <- ExposureFunction(X = dataprep$DistanceShore)

@@

@@

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
  id_track=ggplot()+geom_sf(data=updated_scoresby_sound_utm$geometry,fill="grey")+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=sub_data,aes(px,py),shape=5,size=0.1,col="red")+
  geom_point(data=sub_data,aes(x,y,col=time),size=0.1)+
    geom_path(data=sub_data,mapping=aes(x,y),linewidth=0.1)+
    geom_point(data = sub_data %>% filter(!duplicated(ID)),
             aes(x,y,col=ID), shape = 3, size = 4,col = "red")+
    scale_color_viridis_c()

@@

```{r display Dship_histo, fig.cap="Histograms of distance to ship"}
Dship_histo
```

@@

dataBE1=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE1 =rbind(dataBE1,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime)))
}

dataBE2=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE2 =rbind(dataBE2,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime/2)))
}

dataAE=data.frame()
for (id in unique(dataprep$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataAE =rbind(dataAE,subset(narwhalData, (time >= maxtime) & (time>tagtime)))
}

@@

```{r save preprocessed data}
# Set the path to the directory containing the raw data for narwhals
preprocessed_narwhal_data_path <- here(file.path("Data","preprocessed_data","narwhals"))

write.csv(dataprep,file.path(preprocessed_narwhal_data_path,"allData.csv"))
write.csv(dataBE1,file.path(preprocessed_narwhal_data_path,"DataBE1.csv"))
write.csv(dataBE2,file.path(preprocessed_narwhal_data_path,"DataBE2.csv"))
write.csv(dataAE,file.path(preprocessed_narwhal_data_path,"DataAE.csv"))
```

## Plot ship exposure 12h after tagging

```{r plot real exposure}

cropped_data=dataprep[dataprep$time>12,]
pExpShip=ggplot(cropped_data,aes(time,ExpShip,col=ID))+geom_line(color="black")+
  facet_grid(ID~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Exposure to ship")

```{r}
D_low=0.07
D_up=3
df=cropped_data
df[df$DistanceShore<D_low,"Ishore"]=1/D_low
df[df$DistanceShore>D_up,"Ishore"]=0

pIshore=ggplot(df,aes(time,Ishore,col=ID))+ geom_line(color="black")+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  theme(legend.position="none")+xlab("Time (h)")+ylab("Inverse distance to shore")

ggsave(path=plots_folder,"realIshore_through_time.png", plot=pIshore)
```

```{r}
pIshore
```
